---
title: "ggplot2 quick reference"
output:
  github_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'svg',       # set output to svg
                      warning = FALSE, 
                      message = FALSE,
                      fig.width=8, 
                      fig.height=6,
                      fig.align = 'left')
```

# test123 {.tabset}

## Plots

### Table of contents

1.  [Scatter plots](#scatter)
2.  [Bar plots](#bar)
3.  [Histograms](#hist)
4.  [Box plots](#box)
5.  [Distributions](#dist)
6.  [Time series](#ts)

```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)

# Library for combining ggplots 
library(patchwork)

```

### Scatter plots {#scatter}

You can separate data on `color`, `Shape` and `size`. Position `jitter` and `alpha` can be used to make plots more readable

```{r fig.align='left'}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point(size = 4) +
  xlab("Sepal length") +
  ylab("Sepal width")
```

```{r}
#library(GGally)
#ggpairs(iris, columns = 1:4, aes(color=Species))
```

### Bar plots {#bar}

Bar plots can be used either through `geom_bar()`, which counts the number of instances of each group of x or y, or through `geom_col()` which takes the height/length of the bars.

```{r}
iris %>% 
  filter(Sepal.Length > 5) %>% 
  
  ggplot(aes(x=Species, fill=Species)) +
  geom_bar()
```

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise_all(mean) %>% 
  pivot_longer(-Species, values_to = "mean") %>%
  
  ggplot(aes(x=name, y=mean, fill=Species)) +
  geom_col(position = "dodge2")
```

### Histograms {#hist}

Default position is `"stack"`. Use position `"identity"` for overlapping histograms. Bins/binwidth is determined automatically for this example. Change bins with `bins = n` or `binwidth = c`.

```{r}
ggplot(iris, aes(x=Sepal.Length, fill=Species)) +
  geom_histogram()

```

### Box plots {#box}

```{r}
ggplot(iris, aes(x=Sepal.Length, y=Species, fill=Species)) +
  geom_boxplot()
```

Grouped example of box plots. position defaults to `"dodge2"`.

```{r}
iris %>%
  pivot_longer(cols = -Species) %>%
  ggplot(aes(y=value, x=name, fill=Species)) +
  geom_boxplot() + 
  labs(y = "cm", x="")
```

### Distributions {#dist}

D `"fill"`

`kernel` can be set to one of `"gaussian"`, `"rectangular"`, `"triangular"`, `"epanechnikov"`, `"biweight"`, `"cosine"` or `"optocosine"` with `kernel = "gaussian"` as default.

```{r}
ggplot(iris, aes(x=Sepal.Length, color=Species, fill = Species)) +
  geom_density(alpha=0.6) +
  xlim(bounds=c(4, 8.5))
```

For stacked distribution plots or "ridgeline" plots the library `ggridges` can be used in ggplot syntax. `scale` determines the maximum height of the distribution plots and can be left empty. Plots through this method might need some manual scaling of the y axis through `scale_y_dicrete()` for the desired result.

```{r}
library(ggridges)

ggplot(iris, aes(x=Sepal.Length, y=Species, color=Species, fill=Species, height=after_stat(density))) +
  geom_density_ridges(alpha=0.8, scale=0.9, rel_min_height=0.005, stat="density") +
  theme(legend.position = 'none')  +
  scale_y_discrete(expand = expansion(add = c(0.2, 1))) +
  xlim(c(4,8.5))

```

```{r}
ggplot(iris, aes(x=Sepal.Length, fill=Species, color=Species)) +
  geom_density(show.legend = F, alpha=0.8) + 
  facet_grid(rows=vars(Species)) +
  xlim(c(4,8.5))
```

Borrowed example from [The R Graph Gallery](https://r-graph-gallery.com/294-basic-ridgeline-plot.html#color) with axis flipped. In my opinion, this is a better option than the very similar violin plot when many distributions are to be shown in comparison to each other.

```{r fig.width=10, fig.height=5}

ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = Month, fill = stat(x))) +
  geom_density_ridges_gradient(scale = 0.95, rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]", option = "C") +
  coord_flip()
```

### Time series {#ts}

Using the tsibble and feasts libraries, many A Simple plot of a time series.

```{r}
# Time series libraries
library(tsibble)
library(feasts)

air_pas <-  as_tsibble(AirPassengers)

air_pas %>% 
  autoplot() + 
  ylab("Air passengers (in thousands)") + 
  xlab("") +  
  ggtitle("Monthly international air passengers 1949-1960")
```

Within the feasts/tsibble framework

```{r fig.height=8}
p_seas <- gg_season(air_pas, labels = "right", labels_repel = TRUE)
p_sub <- gg_subseries(air_pas)

p_seas / p_sub #plots combined through patchwork (library) 
```

The same follows for a plot of a time series decomposition. Here STL is used. The trend and seasonal windows are determined automatically for this example.

```{r}
air_pas.stl <- air_pas %>%
  mutate(value = log(value)) %>% # Log series since seasonality seems multiplicative with trend
  model(
    STL(value ~ trend() + season())) # The windows for trend and seasonality are automatically determined for this example
  
air_pas.stl %>%
  components() %>%
  autoplot() + labs(title = "STL decomposition of log(AirPassengers)")

```

A plot for autocorrelation and partial autocorrelation can easily be made from a series. Note that the remainder seems to show some heteroscedasticity and as a result probably isn't stationary.

```{r}
air_pas.stl %>%
  components() %>%
  gg_tsdisplay(y = remainder, plot_type = 'partial')

```

Forecasts with accompanying confidence intervals can be displayed and estimated with `library(fable)`. Set `level` to change the confidence intervals, with `NULL` for no intervals or a list of numbers representing the percentages. Defaults to `levels=c(80,95)`.

```{r}
library(fable)
ts <- tsibbledata::aus_production %>% 
  select(Electricity) 

fit <- ts %>%
  model( ARIMA(Electricity), ETS(Electricity))

fit %>% 
  forecast(h="5 years") %>% 
  autoplot(ts %>% filter(Quarter > yearquarter("1995 Q1")), level=c(80, 95, 99)) + 
  facet_grid(rows=vars(.model))
```

## Interactive charts

Plots from ´ggplot2´ can be turned into interactive plots through the library [plotly](https://github.com/plotly/plotly.py) usually with the ease of "plug and play". Here are a few examples of some interactive charts using the method.

```{r eval=FALSE}
library(plotly)
p <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species, size = Petal.Length )) + 
  geom_point() +
  xlab("Sepal length") +
  ylab("Sepal width")

ggplotly(p)

```

```{r eval=FALSE}
p <- iris %>%
  pivot_longer(cols = -Species) %>%
  ggplot(aes(y=value, x=name, fill=Species)) +
  geom_boxplot(position = position_dodge2(preserve = "single")) + 
  labs(y = "cm", x="") 

ggplotly(p) %>% 
  layout(boxmode = "group")
```


## Theme settings

```{r}
p1 <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point(show.legend = F) +
  xlab("Sepal length") +
  ylab("Sepal width")

(p1 + theme_gray() + ggtitle("Gray")) +
(p1 + theme_minimal()+ ggtitle("Minimal")) +
(p1 + theme_bw()+ ggtitle("Black and white")) +
(p1 + theme_classic()+ ggtitle("Classic")) +
(p1 + theme_dark()+ ggtitle("Dark")) +
(p1 + theme_light()+ ggtitle("Light"))

```

### Colors

Color palettes can be defined through lists default colors supported in ggplot2 or through the colors RGB representation. They can also be named in cases where matches between specific groups and colors are desired.

```{r fig.height=4, fig.width=10}
my_palette1 <- c('purple', 'orange', 'aquamarine3')
my_palette2 <- c('virginica' = '#FF0000', 'setosa' = '#00FF00', 'versicolor' = '#0000FF') # Fully Red, Green, Blue in RGB notation

p1 <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point() +
  xlab("Sepal length") +
  ylab("Sepal width") 

(p1 + scale_color_manual(values = my_palette1)) +
  (p1 + scale_color_manual(values = my_palette2))

```

Here are some examples of color palettes from the [ColorBrewer set](https://colorbrewer2.org). These can easily be used with ggplot through the function `scale_colour_brewer(palette = "palette name")`. Further palettes can be acquired from the R package [Paletteer](https://emilhvitfeldt.github.io/paletteer/) which contains more than 2000 color palettes, including continuous palettes, and functions for simple use with ggplot.

The qualitative colors are useful for distinguishing between different groups of data. For example, different flower species in the iris data set.

```{r}
library(RColorBrewer)
display.brewer.all(type='qual')
```

Some examples of sequential colors, useful for ordered data visualizations such as heat maps.

```{r fig.height=9}
display.brewer.all(type='seq')

```

The diverging colors diverge in two directions from a central color, emphasizing a mid-range value and/or the extremes. Useful for comparisons in relation to a chosen value/group or for loose groupings of ordered data.

```{r}
display.brewer.all(type='div')
```
