```{r}
#| output: false
#| echo: false

library(ggplot2)
```

### Legends

Legends in ggplot are very customizable in position and theme.  

Many of these options are advised to be set on a theme basis rather than individual basis, however, here they are set on an individual basis to show more examples within a small space.

::: panel-tabset
#### Position

-   **guide_legend()** and **guide_colorbar()** are used in conjunction with `guides()` to specify aspects of legend guides ([Legend reference](https://ggplot2.tidyverse.org/reference/guide_legend.html), [Colorbar reference](https://ggplot2.tidyverse.org/reference/scale_colour_continuous.html)). Here are some of the arguments:
    -   **position**: Position of the legend relative to the chart. One of `"top"`, `"right"`, `"bottom"`, `"left"` or `"inside"`
    -   **order** (integer): Order in which the guides are shown.
    -   **ncol/nrow**: Sets number of columns or rows of legend items (only guide_legend)
    -   **direction**: Direction of the legend (`horizontal` or `vertical`)
-   **legend.position**: Default position of the legends

```{r}
#| label: legend_pos
p <- mtcars |> 
  ggplot(aes(x=mpg, y=wt, color=hp, shape=as.factor(gear), size=disp, fill = as.factor(cyl))) +
  geom_point() +
  guides(
    color = guide_colorbar(position = "top", order = 1),
    
    shape = guide_legend(position = "inside", direction = "horizontal"),
    
    size = guide_legend(position = "top", ncol = 2, order = 2),
    
    fill = guide_legend(position = "inside")
  ) + 
  scale_size_binned(range = c(1, 8), n.breaks = 8) +
  theme(legend.position.inside = c(0.8, 0.8))
p
```

#### Justification

-   **legend.position.inside**: Legend placement when legend position is set to inside. (`c(a, b)`, where $a,b \in [0,1]$).
-   **legend.justification**: Justification of legends relative to plot. One of "left", "right", "top", "bottom", "center" or 2-dimensional vector (text or numeric). E.g. `c(0, 1)`, `c("left", "top")`.
    -   can be set individually for each legend position with: **legend.justification.top**, **legend.justification.bottom**, **legend.justification.left**, **legend.justification.right**, **legend.justification.inside**.
-   **legend.box.just**: Justification of legends within the bounding box.
-   **legend.box**: Arrangement of legends within the bounding box. `"horizontal"` or `"vertical"`

Legend bounding box is set with a background and margin here to show its relation to the legends.

```{r}
#| label: legend_just
p  +
  theme(legend.position.inside = NULL,
        legend.justification = "left",
        legend.justification.inside = c("right","top"),
        
        legend.box.background = element_rect(fill="gray"),
        legend.box.margin = unit(2, "mm"),
        legend.box.just = c("center", "bottom"),
        legend.box = "horizontal")
```

#### Size and text

-   **legend.title.position**: Legend title position relative to legend. Can be set to one of `"top"`, `"bottom"`, `"left"`, `"right"`
-   **legend.title**: Options for legend title. Can be set with `element_text()` (see [elements](#elements)).
-   **legend.text**: Options for the key-text. Can be set with `element_text()`.
-   **legend.text.position**: The position of the text relative to the keys. One of `"top"`, `"bottom"`, `"left"`, `"right"`

```{r}
#| label: legend_text
mtcars |> 
  ggplot(aes(x=mpg, y=wt, color=hp, shape=as.factor(gear), size=disp)) +
  geom_point() +
  guides(
    color = guide_colorbar(position = "top",
                           theme = theme(legend.key.width = unit(1, "null"),
                                         legend.title.position = "top",
                                         legend.title = element_text(hjust = .5, size=15))),
    
    shape = guide_legend(theme = theme(legend.text.position = "left")),
    
    size = guide_legend(theme = theme(legend.text.position = "top"))
  ) + 
  labs(color= "Centered and enlarged legend title")
  
```



#### Colors and spacing

Background colors and borders for the components of legends can be set with `element_rect()` (see [element](#element)). 
The theme options also contain many options for spacing and margins, where spacings can be set with `unit()` and margins with `margin()`



For a full list of modifiable elements see [theme reference](https://ggplot2.tidyverse.org/reference/theme.html).

```{r}
#| label: legend_color
mtcars |> 
  ggplot(aes(x=mpg, y=wt, color=hp, shape=as.factor(gear), size=disp)) +
  geom_point() +
  guides(
    size = guide_legend(theme = theme(legend.background = element_rect(fill="orange", color="aquamarine"))),
    
    shape = guide_legend(theme = theme(legend.key = element_rect(fill = "purple"),
                                       legend.key.spacing.y = unit(1, "mm"))),
    
    color = guide_colorbar(position = "top", 
                           theme = theme(legend.frame = element_rect(color = "red")))
  )
```







:::
