
```{r}
#| echo: false
#| output: false
library(ggplot2)
library(tidyr)
```


### Plotly
Plots from ggplot2 can be turned into interactive plots through the library [plotly](https://github.com/plotly/plotly.R), usually with the ease of "plug and play". The resulting charts are equipped with a hover-over tooltip, panning, zooming and selections which can, for example, be used with shiny.

Here are a few examples of some interactive charts using the method.

::: panel-tabset
#### Scatter plot
Variables not present in the chart can be added to the hover-over tooltip by adding them to the ggplot2 aesthetics with a chosen name. For example, label1, label2, etc.

```{r}
#| label: plotly-scatter
library(plotly)
p <- ggplot(iris, aes(color=Species, x=Sepal.Length, y=Sepal.Width, label1=Petal.Width, label2=Petal.Length)) + 
  geom_point() +
  xlab("Sepal length") +
  ylab("Sepal width")

ggplotly(p)

```

#### Box plot

```{r}
#| label: plotly-box
p <- iris |>
  pivot_longer(cols = -Species) |>
  ggplot(aes(y=value, x=name, fill=Species)) +
  geom_boxplot(position = position_dodge2(preserve = "single")) + 
  labs(y = "cm", x="") 

ggplotly(p) |> 
  layout(boxmode = "group")
```

#### Histogram

```{r}
#| label: plotly-hisst
p <- ggplot(iris, aes(x=Sepal.Length, fill=Species)) +
  geom_histogram()

ggplotly(p)
```
:::